---
title: "Cleaning"
---

## Goals of this notebook

The steps we'll take to prepare our data:

- Download the data
- Import it into our notebook
- Clean up data types and columns
- Export the data for next notebook
- Make sure that the updated data that covers the year 2023, aligns with the rest of the data. 

## Setup 

Loading up the libraries: 

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
library(readxl)
```


## Downloading Data 

Here I will download the TCEQ data that I obtained from a public records request. The data goes over the counties or jurisdictions affected by water boil notices, start and end date of a water boil notice and the nature of the water boil notice. 

More information about the data can be found on the index page of this notebook. 

```{r}
download.file(
  "https://utexas-my.sharepoint.com/:x:/g/personal/mp48828_my_utexas_edu/EZZunjkd_7dLmH6EduWTVu4BFVPcKRpeGRoLIYyfYyrG7w?e=8pRzRN",
  "data-raw/old_water_boil.xlsx",
  mode = "wb"
)
```
```{r}
read_excel("data-raw/old_water_boil.xlsx")
```




